Weighted Model Counting with Conditional Measures

### 2020-07-20 paper.tex
Move things around

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

\title{Weighted Model Counting with Conditional Probabilities}
\author{Paulius Dilkas}

@@

\paragraph{Legend.}
\begin{description}
\item[F] Feedback (round 1).
\item[F2] Feedback (round 2).
\item[FP] Feedback from the review panel.
\item[\textbullet] My own idea.
\end{description}

\paragraph{The big TODO list.}
\begin{itemize}
\item Look into ways to theoretically justify the performance benefits. Some
  suggestions:
  \begin{itemize}
  \item the number of variables,
  \item the number of clauses/CPTs/lines in CPTs,
  \item the proportion of edges that exist,
  \item average degree,
  \item average number of clauses that a variable is in,
  \item the number of clusters in the linear/tree-like decomposition.
  \end{itemize}
\item Have an example of how the ADDs function in this situation. If not for the
  paper, then at least for slides. Use the framework to check its correctness.
\item[F] Give a concrete example of something impossible to represent using WMC.
\item Turn the saved PDFs into citations.
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow. To do this, we appeal to algebraic constructions to
  define the main concepts for introducing measures on Boolean algebras.
\item[F] Can you say something here about factorized vs non-factorized
  weight function definitions? That is, factorized is when $w$ maps literals to
  $R_{\ge 0}$, non-factorized is when $w$ maps models to $R_{\ge 0}$ and
  \begin{itemize}
  \item come up with nice example when non-factorized weights are intuitive;
  \item clarify that the factorized definition have is w.r.t. models, in case some
    one gets confused. [It doesn't have to be, if the BA is not free---P.]
  \end{itemize}
\item[F2] The paper at this stage is very technical---the danger is that WMC/SRL
  people may not be able to follow it and so would be hard to get accepted.
  Without clear target audience, [???] get work accepted. My main high level
  suggestion is that let us tease apart what you have and see if a story
  emerges. That is, let us attempt to write a paper {\bf with examples} and
  see whether with significant motivation, we have a story emerging. Below,
  sample text needed to adequately motivate [???] for WMC/SRL community.
\item[FP] Clear up questions about independence between literals.
\item[F2] Preliminaries: explain with examples: models $=$ elements [atoms] of
  algebra.
\item WMC as a Measure
  \begin{itemize}
  \item[F2] Need to explain how WMC and NWMC connects to standard definitions of
    WMC and $\Pr(\phi \mid e) = \WMC(\phi \land e)/\WMC(e)$.
  \item[F2] You need to explain what precisely these mean in logic and models
    and weight functions are usually defined and understood.
  \item Be careful about mentioning ideals, filters, and quotients.
  \end{itemize}
\item What Measures are WMC-Definable: proofs need to be updated and
  propositions could be phrased in a better way, but the gist should be the
  same.
\item Extending the Algebra.
  \begin{itemize}
  \item[F2] If you prove (b) above, you could motivate why weights on literals
    is attractive and whether there is a way to augment the expressivity of WMC
    while still maintaining literal level weights. Hence this action.
  \item[F2] You need to explain the significance of this result.
  \end{itemize}
\item Make up my mind about $a, b$ vs. $x, y$ and stick to it (maybe $x, y$?).
\item Terminology: `with generating set $S$' $\to$ `over $S$'.
\item Perhaps reorder the section of preliminaries into paragraphs, i.e., a
  paragraph for order, for homomorphisms, etc. This would take up less space.
\item Notation: if $L$ denotes literals, then it doesn't denote a generating
  set.
\item Prove that we have an algebra of functions with two (?) additional
  operations. Or maybe just cite the original paper for this.
\item The function $\phi$ created by the algorithm can be seen as a measure on
  the BA $2^U$. Can I at least show that this BA has fewer elements as a measure
  of compactness w.r.t. the measure?
\item Note that we're going to use logical notation for BAs. Give an example of
  the two notations.
\item Could clarify why $[\lambda_{X=0}] = \overline{[\lambda_{X=1}]}$. This
  follows from the definition. Or should I avoid the overline notation?
\item After the experiments are finished, note the processor, memory per thread,
  and add the following acknowledgment. This work has made use of the resources
  provided by the Edinburgh Compute and Data Facility (ECDF)
  (\url{http://www.ecdf.ed.ac.uk/}).
\end{itemize}

\paragraph{Experimental setup.}
\begin{itemize}
\item
  Algorithms\footnote{\url{http://beyondnp.org/pages/solvers/model-counters-exact/}}
  \begin{itemize}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} (rediscovered the
    multiplicativity of BAs in different words) (with optimal settings)
  \item Cachet \cite{DBLP:conf/sat/SangBBKP04}
  \item c2d \cite{DBLP:conf/ecai/Darwiche04}
  \item d4 \cite{DBLP:conf/ijcai/LagniezM17} (closed source, boo!)
  \item miniC2D  \cite{DBLP:conf/ijcai/OztokD15}
  \end{itemize}
\item Encodings:
  \begin{itemize}
  \item \texttt{-d02} \cite{DBLP:conf/kr/Darwiche02}
  \item \texttt{-sbk05} \cite{DBLP:conf/aaai/SangBK05}
  \item \texttt{-cd05} \cite{DBLP:conf/ijcai/ChaviraD05}
  \item \texttt{-cd06} \cite{DBLP:conf/sat/ChaviraD06} (supposed to be the best)
  \item mine
  \end{itemize}
  All except mine are from Ace
  3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} and should be compiled
  with \texttt{-encodeOnly} (i.e., don't compile the CNF into an AC) and
  \texttt{-noEclause} (i.e., only use standard syntax) flags.
\item Datasets\footnote{There might be more at
    \url{https://meelgroup.github.io/}.}
  \begin{itemize}
  \item binary Bayesian networks from Sang et
    al.\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}
    \cite{DBLP:conf/aaai/SangBK05}
    \begin{itemize}
    \item Grid (networks) (ratio 75 means that 75\% of the nodes are
      deterministic),
    \item Plan recognition (problems),
    \item Deterministic quick medical reference (what do the numbers mean? the
      README doesn't say).
    \end{itemize}
  \item Bayesian networks available with Ace
    \begin{itemize}
    \item \texttt{2004-pgm} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \item \texttt{2005-ijcai} \cite{DBLP:conf/ijcai/ChaviraD05}. The Genie/Smile
      files have their own citation data that I should probably extract. This is
      the only dataset that has some non-binary networks.
    \item \texttt{2006-ijar} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \end{itemize}
  \item ProbLog \cite{DBLP:conf/uai/FierensBTGR11}
  \item probabilistic programs \cite{DBLP:journals/corr/abs-2005-09089}
  \end{itemize}
\end{itemize}

\paragraph{Contributions.}
\begin{itemize}
\item WMC defines a measure over a BA.
\item WMC with weights on literals imposes an independence assumption. (Measures
  are `slightly' more expressive than WMC with weights on models because they
  apply to non-atomic BAs.)
\item A BA can be augmented with new literals in order to support any measure.
\item (Maybe) a lower bound on the number of new literals needed in order to
  support any measure.
\item Alternatively, one can use coproducts and pushouts to define a BA with
  precisely the right independence and conditional independence conditions.
  (This requires a relaxed version of WMC.)
\item This results in a smaller problem for WMC algorithms (w.r.t. both the
  number of literals and the length of the theory) and is optimal for, e.g.,
  Bayesian networks.
\item (Maybe) this results in faster inference (?)
\end{itemize}

\paragraph{Notable previous/related work.}
\begin{itemize}
\item Hailperin's approach to probability logic
  \cite{DBLP:journals/ndjfl/Hailperin84}
\item Nilsson's (somewhat successful) probabilistic logic
  \cite{DBLP:journals/ai/Nilsson86}
\item Logical induction: a big paper with a good overview of previous attempts
  to assign probabilities to logical sentences in a sensible way
  \cite{DBLP:journals/eccc/GarrabrantBCST16}
\item Measures on Boolean algebras
  \begin{itemize}
  \item On possibility and probability measures in finite Boolean algebras
    \cite{DBLP:journals/soco/CastineiraCT02}
  \item Representation of conditional probability measures
    \cite{krauss1968representation}
  \end{itemize}
\end{itemize}

\paragraph{Notes.}
\begin{itemize}
\item Thesis: many important computational problems are solved by encoding them
  as WMC. But if the weight function is extended to `conditional probabilities',
  the problem becomes easier because current approaches need workarounds in
  order to be independent.
\item Shorter thesis: many important problems are encoded as WMC, but they can
  be encoded in a better way if we allow for conditional probabilities.
\item We extend the Gaifman graph to add edges when two variables occur in the
  same CPT (e.g., including the edge from $A$ to $B$ when the CPT is $\Pr(A \mid
  B)$).
\item Extra benefit: one does not need to come up with a way to turn some probability
distribution to into a fully independent one.
\item Trivial CPTs such as $\Pr(A \mid B) = \Pr(\neg A \mid B) = 0.5$, the ADDs
  of which simplify to a single number, are put in the first cluster.
\item Observation: by inspecting the BN, we could identify CPTs that could be
  completely ignored, but maybe good heuristics take care of that anyway.
\item Observation: ADDs have a restrict command. I'm not going to use it, but it
  could be nice.
\item Alternative scenario to consider for future work: \#\SAT{} solver
  generates solutions which are then used to restrict the ADDs that calculate
  probabilities.
\item Important future work: replacing ADDs with
  AADDs\footnote{\url{https://github.com/ssanner/dd-inference}} is likely to
  bring performance benefits. Other extensions:
  \begin{itemize}
  \item FOADDs can represent first order statements;
  \item XADDs can replace WMI for continuous variables;
  \item ADDs with intervals can do approximations.
  \end{itemize}
\item cd06 is supposed to be the fastest, but seems to be one of the slowest
  (along with cd05). Perhaps the best encoding for arithmetic circuits has very
  little to do with the best encoding for CNF WMC solvers?
\item When the Bayesian network has an evidence file, we compute the probability
  of evidence. Otherwise, let $X$ denote the last-mentioned node in the Bayesian
  network. If $\mathsf{true}$ is a valid value of $X$, we compute the marginal
  probability of $X = \mathsf{true}$. Otherwise, we pick the first value of $X$
  and calculate its marginal probability. This applies to the Grid data set (as
  intended) and also to two instances of Plan Reconstruction and roughly half of
  the instances from 2004-PGM that have empty evidence files.
\item Bayesian networks are often solved in a compile once, query many times
  fashion. This can be achieved using ADDMC by selecting a subset $S$ of
  variable we may want to query over and running ADDMC while excluding $S$ from
  variable elimination/projection/$\exists$.
\item \texttt{cd05} relaxes the encoding so much that extra models become
  possible. They are supposed to be filtered out by the algorithm, but mine
  can't do that because it doesn't deal with models. Same for \texttt{cd06}
  because it's based on \texttt{cd05}.
\item For ground ProbLog, we can encode a program
\begin{verbatim}
p :: a :- b
q :: a :- c
\end{verbatim}
  into $P(a \mid b)=p$, $P(a \mid c)=q$ instead of having clauses $b \Rightarrow
  a$, $c \Rightarrow a$. Some logical structure is likely to remain.
\item Potential criticism may be that this is a step backwards and doesn't allow
  us to use SAT-based techniques for probabilistic inference. However, they can
  still be used for the 'theory+query' part.
\item We don't compare 'compile times' because our encoding time is linear, so
  we would easily beat everyone else.
\item The claim behind this paper is that allowing for conditional probabilities
  in the context of weighted model counting seems to be a good idea. Bayesian
  networks and ADDMC are only particular examples. This should also work with
  Cachet.
\item For the ProbLog $\to$ WMC conversion, check out this
  guy.\footnote{\url{https://users.ics.aalto.fi/ttj/}}
\item Zero-probability weights and one-probability weights can be interpreted as
  logical clauses. This doesn't affect ADDMC but could be useful for other
  solvers.
\item Filtering out ADDs that have nothing to do with the answer helps
  tremendously, but I'm not sure if I should do that. Could a heuristic do the
  same thing?
\item For each row in each CPT, we order values by decreasing probabilities.
  Need to check if this makes the solver more accurate.
\item In the encoding:
  \begin{itemize}
  \item We assume that the first literal after `w' is positive.
  \item The last two numbers are the positive and the negative probabilities,
    respectively---sometimes they add to one, and sometimes the negative
    probability is one, regardless of the value of the first probability.
  \end{itemize}
\item My encoding for non-binary BNs should still be faster because it doesn't
  have those `equivalence' clauses.
\item Guess: what makes the problem easier is:
  \begin{enumerate}
  \item short clauses,
  \item low number of clauses.
  \end{enumerate}
  This explains why my first version of encoding non-binary CPTs was too slow.
\item With all encodings (including mine non-binary encoding), they live on a
  measure Boolean algebra where the measure is not probabilistic (so $\Pr(\neg
  A) \ne 1 - \Pr(A)$).
\item We assume that all variables in the Bayesian network have at least two
  values.
\item Remark: for any function $A\colon 2^X \to \mathbb{R}_{\ge 0}$, $A +
  \overline{A} = 1$.
\end{itemize}

@@

@@

\section{WMC as a Measure}

@@

\section{What Measures Are WMC-Definable?}

@@

@@

\section{Representing Independence and Conditional Independence}

\subsection{Preliminaries}

\begin{definition}
  Given a BA $\mathbf{A}$, a \emph{subalgebra} is a subset $\mathbf{B} \subseteq
  \mathbf{A}$ that, together with the operations, zero, and one of $\mathbf{A}$,
  is a BA.
\end{definition}

\begin{definition}[\cite{givant2008introduction}]
  Let $\mathbf{A},$ $\mathbf{B}$, and $\mathbf{C}$ be BAs such that $\mathbf{B}$
  is a subalgebra of $\mathbf{A}$. Let $f\colon \mathbf{A} \to \mathbf{C}$ and
  $g\colon \mathbf{B} \to \mathbf{C}$ be homomorphisms. Then $f$ is an
  \emph{extension} of $g$ if $f(x) = g(x)$ for all $x \in \mathbf{B}$. If $f$ is
  an extension of each member of a family $\{ g_i \}_{i \in I}$ of
  homomorphisms, then $f$ is called a \emph{common extension} of $\{ g_i \}_{i
    \in I}$.
\end{definition}

\begin{definition}[\cite{givant2008introduction}]
  Let $\{ \mathbf{A}_i \}_{i \in I}$ be a family of subalgebras of a BA
  $\mathbf{A}$. If for any BA $\mathbf{B}$ with a family of homomorphisms $\{
  f_i\colon \mathbf{A}_i \to \mathbf{B} \}_{i \in I}$ there exists a unique
  common extension of $\{ f_i\colon \mathbf{A}_i \to \mathbf{B} \}_{i \in I}$
  ($f\colon \mathbf{A} \to \mathbf{B}$ in the diagram),
  \[
    \begin{tikzcd}
      \mathbf{A}_i \ar[r,hookrightarrow] \arrow{rd}[swap]{f_i} & \mathbf{A}
      \arrow[d,dashed,"f"] \\
      & \mathbf{B}
    \end{tikzcd}
  \]
  then $\mathbf{A}$ is the \emph{internal sum}\footnote{A slightly more general
    version of this definition is also known as the free product, the Boolean
    product, and the coproduct in the category of BAs
    \cite{givant2008introduction,koppelberg1989handbook,sikorski1969boolean}.}
  of $\{\mathbf{A}_i \}_{i \in I}$. We will denote it as $\bigoplus_{i \in I}
  \mathbf{A}_i$.
\end{definition}

\begin{proposition}[\cite{sikorski1969boolean}]
  Let $\mathbf{A}$ be the internal sum of a family of BAs $\{ \mathbf{A}_i \}_{i
    \in I}$, and let $\{m_i\colon \mathbf{A}_i \to \mathbb{R}_{\ge 0} \}_{i \in
    I}$ be a family of measures. Then there exists a unique measure $m\colon
  \mathbf{A} \to \mathbb{R}_{\ge 0}$ such that, for any finite subset $J
  \subseteq I$ and family of elements $\{ x_j \in \mathbf{A}_j \}_{j \in J}$,
  \[
    m \left( \bigwedge_{j \in J} x_j \right) = \prod_{j \in J} m_j(x_j).
  \]
\end{proposition}

\begin{definition}[\cite{koppelberg1989handbook}]
  Let $\mathbf{A}$ be a BA. Let $\mathbf{B}$ be a subalgebra of $\mathbf{A}$,
  and let $\{ \mathbf{A}_i\}_{i \in I}$ be a family of subalgebras of $\mathbf{A}$
  such that $\mathbf{A}_i \cap \mathbf{A}_j = \mathbf{B}$ for all $i \ne j$ in
  $I$. Let $\{ \iota_i \colon \mathbf{B} \to \mathbf{A}_i \}$ be a family of
  inclusion homomorphisms. Then $\mathbf{A}$ is the \emph{amalgamated free
    product\footnote{Also known as a (wide) pushout in the category of BAs.} of
    $\{\mathbf{A}_{i} \}_{i \in I}$ over $\mathbf{B}$} if, for any Boolean
  algebra $\mathbf{C}$ with a family of homomorphisms $\{ f_i\colon \mathbf{A}_i
  \to \mathbf{C} \}_{i \in I}$ such that $f_i \circ \iota_i = f_j \circ \iota_j$
  for all $i, j \in I$, there is a unique homomorphism $f\colon \mathbf{A} \to
  \mathbf{C}$ such that the triangle in
  \[
    \begin{tikzcd}
      \mathbf{B} \ar[r,hookrightarrow,"\iota_i"] & \mathbf{A}_i
      \ar[r,hookrightarrow] \ar[d,swap,"f_i"] & \mathbf{A} \ar[ld,dashed,"f"] \\
      & \mathbf{C}
    \end{tikzcd}
  \]
  commutes for all $i \in I$. We will denote this product as
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathbf{B}\\ i \in I}} \mathbf{A}_i.
  \]
\end{definition}

\subsection{New Results}

\todo[inline]{Make sure that this is enough to guarantee a pushout.}
\begin{theorem}
  Let $\{ S_i \}_{i=0}^n$ be a finite set of finite sets for some $n > 1$ such
  that for all distinct positive integers $i$ and $j$, $S_i \cap S_j = S_0$, and
  let
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathcal{F}(S_0)\\ 1 \le i \le n}}
    \mathcal{F}(S_i).
  \]
  Let $(m_i\colon \mathcal{F}(S_i) \to \mathbb{R}_{\ge 0})_{i=1}^n$ be arbitrary
  measures. Then there is a unique measure $m\colon \mathbf{A} \to
  \mathbb{R}_{\ge 0}$ such that, for any element $b \in \mathcal{F}(S_0)$,
  subset $J \subseteq \{ 1, 2, \dots, n \}$, and elements $\{ a_j \in
  \mathcal{F}(S_j \setminus S_0) \}_{j \in J}$,
  \[
    m \left(b \land \bigwedge_{j \in J} a_j \right) = \prod_{j \in J} m_j(b
    \land a_j).
  \]
\end{theorem}

\begin{theorem}
  The number of weights needed to encode a Bayesian network using
  coproducts and pushouts is equal to the number of entries in the tables of the
  network (and the resulting theory is shorter).
\end{theorem}

\begin{theorem}[Pushouts of free BAs are free]
  Let
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathbf{B}\\ i \in I}} \mathbf{A}_i
  \]
  be an amalgamated free product such that $\{ \mathbf{A}_{i} \}_{i \in I}$ are
  free BAs with $\{ S_i \}_{i \in I}$ as their respective sets of generators.
  Let $S = \bigcup_{i \in I} S_i$. Then $\mathbf{A}$ is a free BA with
  generating set $S$.
\end{theorem}
\begin{proof}
  Suppose we have a map from $S$ to an arbitrary BA $\mathbf{C}$, as in
  \[
    \begin{tikzcd}
      S_i \ar[r,hookrightarrow] \ar[d,hookrightarrow] & S \ar[d,hookrightarrow]
      \ar[ddr,bend left] & \\
      \mathbf{A}_i \ar[r,hookrightarrow] \ar[rrd,dashed,bend right] & \mathbf{A}
      \ar[rd,dashed]& \\
      & & \mathbf{C}.
    \end{tikzcd}
  \]
  We want to show that there exists a unique homomorphism $\mathbf{A} \to
  \mathbf{C}$. For all $i \in I$, from $S_i \hookrightarrow S$ and $S \to
  \mathbf{C}$ we get a map $S_i \to \mathbf{C}$, so---by the definition of a
  free BA---there is a unique homomorphism $\mathbf{A}_i \to \mathbf{C}$.
  Furthermore, a family of homomorphisms $\{ \mathbf{A}_i \to \mathbf{C} \}_{i
    \in I}$ uniquely determine a homomorphism $\mathbf{A} \to \mathbf{C}$ by the
  universal mapping property of a (wide) pushout. Thus $\mathbf{A}$ is a free BA
  with generating set $S$.
\end{proof}

\begin{corollary}
  Similarly, coproducts of free BAs are free.
\end{corollary}

\section{Representation of conditional probability tables}

The probability of a variable with $n$ values $x_1, \dots, x_n$ (whether
conditioned on something or not) can be represented with an ADD of size
$\Theta(n^2)$ by representing each of $\Pr(x_1), \Pr(x_2 \mid x_1), \dots,
\Pr(x_n \mid x_1, \dots, x_{n-1})$ by a separate ADD. For example, if $n=3$, the
last of the three ADDs would look like this
\[
  \begin{tikzpicture}
    \node[ellipse,draw] (x3) {$x_3$};
    \node[ellipse,draw,below left = of x3] (x21) {$x_2$};
    \node[ellipse,draw,below right = of x3] (x22) {$x_2$};
    \node[ellipse,draw,below = of x21] (x11) {$x_1$};
    \node[ellipse,draw,below = of x22] (x12) {$x_1$};
    \node[rectangle,draw,below = of x11] (n2) {$0.7$};
    \node[rectangle,draw,left = of n2] (n1) {$0$};
    \node[rectangle,draw,below = of x12] (n3) {$0.3$};
    \node[rectangle,draw,right = of n3] (n4) {$1$};

    \draw (x3) edge (x21);
    \draw (x21) edge (n1);
    \draw (x11) edge (n1);
    \draw[dashed] (x21) edge (x11);
    \draw[dashed] (x11) edge (n2);

    \draw[dashed] (x3) edge (x22);
    \draw[dashed] (x22) edge (x12);
    \draw[dashed] (x12) edge (n3);
    \draw (x22) edge (n4);
    \draw (x12) edge (n4);
  \end{tikzpicture}
\]
Such an ADD for $x_n$ has $2n+3$ nodes and $4n-2$ edges. The same information
can be represented in textual form as
\begin{align*}
  &\texttt{w 3 1 0} \\
  &\texttt{w 3 2 0} \\
  &\texttt{w 3 -1 -2 0.7}
\end{align*}
The encoding weights depend on the probabilities in the CPT as follows. For each
value $x_n$, if $\Pr(X = x_n) = 0$, then $w(x_n) = 0$. Otherwise, let $I_n = \{ 1
\le i \le n - 1 \mid \Pr(X = x_i) \ne 0 \}$. Then
\[
  w \left( x_n \;\middle|\; \bigwedge_{i \in I_n} \neg x_i \right) = \Pr(X =
  x_n) \prod_{i \in I_n} \frac{1}{1 - w \left( x_i \;\middle|\; \bigwedge_{j \in
        I_i} \neg x_j \right)}.
\]
All of these weights/probabilities may have other conditions that come from the
structure of the Bayesian network.

\section{Encoding Bayesian networks}

@@

@@

@@

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

\title{Weighted Model Counting with Conditional Measures}
\author{Paulius Dilkas}

@@

\begin{itemize}
\item Thesis: many important computational problems are solved by encoding them
  as WMC. But if the weight function is extended to `conditional probabilities',
  the problem becomes easier because current approaches need workarounds in
  order to be independent.
\item Shorter thesis: many important problems are encoded as WMC, but they can
  be encoded in a better way if we allow for conditional probabilities.
\item The claim behind this paper is that allowing for conditional probabilities
  in the context of weighted model counting seems to be a good idea. Bayesian
  networks and ADDMC are only particular examples. This should also work with
  Cachet.
\item Potential criticism may be that this is a step backwards and doesn't allow
  us to use SAT-based techniques for probabilistic inference. However, they can
  still be used for the 'theory+query' part.
  \begin{itemize}
  \item Zero-probability weights and one-probability weights can be interpreted
    as logical clauses. This doesn't affect ADDMC but could be useful for other
    solvers.
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow. To do this, we appeal to algebraic constructions to
  define the main concepts for introducing measures on Boolean algebras.
\item
  Algorithms\footnote{\url{http://beyondnp.org/pages/solvers/model-counters-exact/}}
  \begin{itemize}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} (rediscovered the
    multiplicativity of BAs in different words) (with optimal settings)
  \item Cachet \cite{DBLP:conf/sat/SangBBKP04}
  \item c2d \cite{DBLP:conf/ecai/Darwiche04}
  \item d4 \cite{DBLP:conf/ijcai/LagniezM17} (closed source, boo!)
  \item miniC2D  \cite{DBLP:conf/ijcai/OztokD15}
  \end{itemize}
\item Contributions
  \begin{itemize}
  \item WMC defines a measure over a BA.
  \item WMC with weights on literals imposes an independence assumption.
    (Measures are `slightly' more expressive than WMC with weights on models
    because they apply to non-atomic BAs.)
  \item A BA can be augmented with new literals in order to support any measure.
  \item (Maybe) a lower bound on the number of new literals needed in order to
    support any measure.
  \item This results in a smaller problem for WMC algorithms (w.r.t. both the
    number of literals and the length of the theory) and thus faster inference.
  \end{itemize}
\item Notable previous/related work
  \begin{itemize}
  \item Hailperin's approach to probability logic
    \cite{DBLP:journals/ndjfl/Hailperin84}
  \item Nilsson's (somewhat successful) probabilistic logic
    \cite{DBLP:journals/ai/Nilsson86,DBLP:journals/ai/Nilsson93}
  \item Logical induction: a big paper with a good overview of previous attempts
    to assign probabilities to logical sentences in a sensible way
    \cite{DBLP:journals/eccc/GarrabrantBCST16}
  \item Measures on Boolean algebras
    \begin{itemize}
    \item On possibility and probability measures in finite Boolean algebras
      \cite{DBLP:journals/soco/CastineiraCT02}
    \item Representation of conditional probability measures
      \cite{krauss1968representation}
    \end{itemize}
  \end{itemize}
\item[F2] The paper at this stage is very technical---the danger is that WMC/SRL
  people may not be able to follow it and so would be hard to get accepted.
  Without clear target audience, [???] get work accepted. My main high level
  suggestion is that let us tease apart what you have and see if a story
  emerges. That is, let us attempt to write a paper {\bf with examples} and
  see whether with significant motivation, we have a story emerging. Below,
  sample text needed to adequately motivate [???] for WMC/SRL community.
\end{itemize}

@@

\begin{itemize}
\item[F2] Explain with examples: models $=$ elements [atoms] of algebra.
\item Perhaps reorder the section of preliminaries into paragraphs, i.e., a
  paragraph for order, for homomorphisms, etc. This would take up less space.
\item Make up my mind about $a, b$ vs. $x, y$ and stick to it (maybe $x, y$?).
\item Terminology: `with generating set $S$' $\to$ `over $S$'.
\item Notation: if $L$ denotes literals, then it doesn't denote a generating
  set. Literals should be $U$.
\end{itemize}

@@

\section{Weighted Model Counting as a Measure}

\begin{itemize}
\item Any function $2^U \to \mathbb{R}_{\ge 0}$ induces a measure on
  $2^{2^U}$---the full Boolean algebra. This measure is not necessarily
  probabilistic.
\item[F2] Need to explain how WMC and NWMC connects to standard definitions of
  WMC and $\Pr(\phi \mid e) = \WMC(\phi \land e)/\WMC(e)$.
\item[F2] You need to explain what precisely these mean in logic and models
  and weight functions are usually defined and understood.
\item Be careful about mentioning ideals, filters, and quotients.
\item Note that we're going to use logical notation for BAs. Give an example of
  the two notations.
\end{itemize}

@@

\section{Limitations of Literal-Based Weighted Model Counting}

\begin{itemize}
\item[F] Give a concrete example of something impossible to represent using WMC.
\item[F] Can you say something here about factorized vs non-factorized
  weight function definitions? That is, factorized is when $w$ maps literals to
  $R_{\ge 0}$, non-factorized is when $w$ maps models to $R_{\ge 0}$ and
  \begin{itemize}
  \item come up with nice example when non-factorized weights are intuitive;
  \item clarify that the factorized definition have is w.r.t. models, in case some
    one gets confused. [It doesn't have to be, if the BA is not free---P.]
  \end{itemize}
\item What Measures are WMC-Definable: proofs need to be updated and
  propositions could be phrased in a better way, but the gist should be the
  same.
\end{itemize}

@@

\begin{itemize}
\item[F2] If you prove (b) above, you could motivate why weights on literals
  is attractive and whether there is a way to augment the expressivity of WMC
  while still maintaining literal level weights. Hence this action.
\item[F2] You need to explain the significance of this result.
\end{itemize}

@@

\section{Previous Work}

\subsection{Bayesian Network Encodings}

\begin{itemize}
\item All encodings (including my non-binary encoding) live on a measure Boolean
  algebra where the measure is not probabilistic (so $\Pr(\neg A) \ne 1 -
  \Pr(A)$).
\item \texttt{cd05} relaxes the encoding so much that extra models become
  possible. They are supposed to be filtered out by the algorithm, but mine
  can't do that because it doesn't deal with models. Same for \texttt{cd06}
  because it's based on \texttt{cd05}.
\item \texttt{sbk05} uses my trick with dividing probabilities. That could
  explain small inaccuracies in its answers.
\item Encodings:
  \begin{itemize}
  \item \texttt{d02} \cite{DBLP:conf/kr/Darwiche02}
  \item \texttt{sbk05} \cite{DBLP:conf/aaai/SangBK05}
  \item \texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05}
  \item \texttt{cd06} \cite{DBLP:conf/sat/ChaviraD06} (supposed to be the best)
  \item \texttt{db20} (mine)
  \end{itemize}
\end{itemize}

\subsection{Algebraic Decision Diagrams and ADDMC}

\section{The Algebra of Functions on Boolean Algebras}

\begin{itemize}
\item Could clarify why $[\lambda_{X=0}] = \overline{[\lambda_{X=1}]}$. This
  follows from the definition. Or should I avoid the overline notation?
\item Prove that we have an algebra of functions with two (?) additional
  operations. Or maybe just cite the original paper for this.
\end{itemize}

@@

\begin{remark}
  For any function $A\colon 2^X \to \mathbb{R}_{\ge 0}$, $A + \overline{A} = 1$.
\end{remark}

@@

\section{Encoding Bayesian Networks Using Conditional Measures}

\begin{itemize}
\item We assume that the first literal after `w' is positive.
\item The last two numbers are the positive and the negative probabilities,
  respectively---sometimes they add to one, and sometimes the negative
  probability is one, regardless of the value of the first probability.
\item We assume that all variables in the Bayesian network have at least two
  values.
\item Have an example of how the ADDs function in this situation. If not for the
  paper, then at least for slides. Use the framework to check its correctness.
\item The function $\phi$ created by the algorithm can be seen as a measure on
  the BA $2^U$.
\item We extend the Gaifman graph to add edges when two variables occur in the
  same CPT (e.g., including the edge from $A$ to $B$ when the CPT is $\Pr(A \mid
  B)$).
\end{itemize}

@@

\section{Experimental Comparison}

\begin{itemize}
\item We don't compare 'compile times' because our encoding time is linear, so
  we would easily beat everyone else.
\item When the Bayesian network has an evidence file, we compute the probability
  of evidence. Otherwise, let $X$ denote the last-mentioned node in the Bayesian
  network. If $\mathsf{true}$ is a valid value of $X$, we compute the marginal
  probability of $X = \mathsf{true}$. Otherwise, we pick the first value of $X$
  and calculate its marginal probability. This applies to the Grid data set (as
  intended) and also to two instances of Plan Reconstruction and roughly half of
  the instances from 2004-PGM that have empty evidence files.
\item After the experiments are finished, note the processor, memory per thread,
  and add the following acknowledgment.
\item All other encodings are implemented in
  Ace~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} and should be
  compiled with \texttt{-encodeOnly} (i.e., don't compile the CNF into an AC)
  and \texttt{-noEclause} (i.e., only use standard syntax) flags.
\item Datasets
  \begin{itemize}
  \item binary Bayesian networks from Sang et
    al.\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}
    \cite{DBLP:conf/aaai/SangBK05}
    \begin{itemize}
    \item Grid (networks) (ratio 75 means that 75\% of the nodes are
      deterministic),
    \item Plan recognition (problems),
    \item Deterministic quick medical reference (what do the numbers mean? the
      README doesn't say).
    \end{itemize}
  \item Bayesian networks available with Ace
    \begin{itemize}
    \item \texttt{2004-pgm} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \item \texttt{2005-ijcai} \cite{DBLP:conf/ijcai/ChaviraD05}. The Genie/Smile
      files have their own citation data that I should probably extract. This is
      the only dataset that has some non-binary networks.
    \item \texttt{2006-ijar} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \end{itemize}
  \end{itemize}
\end{itemize}

\section{Explaining The Performance Benefits}
